---
title: "Report_2"
author: "Antonina Martynenko"
date: '2022-06-24'
output: html_document
---


```{css style settings, echo = FALSE}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 13px;
    border-left: 5px solid #eee;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidytext)
library(kableExtra)

library(wesanderson)
library(pals)
theme_set(theme_minimal())

setwd("~/Documents/est_np/")

# loading data
meta <- read.csv("data/meta_rus_lang.csv")
meta_1880 <- meta %>% 
  filter(year.x < 1891 & year.x > 1879)

meta_1917 <- meta %>% 
  filter(year.x == 1917)

rm(meta)

beta_1880 <- read.csv("data/models/beta_1880_20.csv")
gamma_1880 <- read.csv("data/models/gamma_1880_20.csv")
```

## Exploration with topic modeling

The simple methodological idea beside the following experiments is whether we can use topic modeling on historical OCR-ed newspaper data to spot some historical events happened during a given period. The two periods and problems explored below are: I - Russification in Baltics in 1880s, II - Reactions to two Russian revolutions in 1917.

### Part I 
#### Russification: Switch from German to Russian in local newspapers of Gouvernorates of Estonia and Livonia

This part of report explores the switch from (mostly) German to Russian language of the official newspapers (rus. *Эстляндские губернские ведомости* / ger. *Estländische Gouvernements-Zeitung* and rus. *Лифляндские губернские ведомости* / ger. *Livländische Gouvernements-Zeitung*). As shown in the first report, the digitized newspapers data shows that between 1885 and 1886 the main newspapers of the two Governorates both ended to use German and started to publish texts almost exclusively in Russian. This issue will be studied below using a subset of newspaper pages published in 1880s.  

However, it should be noted that the content of the both newspapers never was in a single language and proportion of German, Russian and Estonian texts can be found for the whole period under consideration.

```{r, echo = FALSE, include = TRUE, fig.width = 9, fig.height=4}
meta_1880 %>% 
  filter(lang_cln %in% c("german", "russian", "estonian")) %>% 
  mutate(title_cln = str_remove_all(docid, "\\d+")) %>% 
  group_by(year.x, title_cln, lang_cln) %>% 
  count() %>% 
  ggplot(aes(x = as.character(year.x), y = n, fill = lang_cln)) + geom_col() + 
  scale_fill_manual(values = c(
    wes_palette("Rushmore1")[5], wes_palette("Rushmore1")[3], wes_palette("Rushmore1")[4])) + 
  facet_wrap(~title_cln, scales = "free") + 
  labs(x = "Year", 
       y = "Number of pages", 
       fill = "Language")
```

The change from predominantly German-pages to Russian happened abruptly in Estonian Gouvernorate's newspaper (ekmteataja) in 1886 and a bit more smoothly at the same time in the Livonian Gouvernorate's one (livzeitung). Number of pages in Estonian stays low in 1880s.
  
For the further experiment a random sample of 3000 pages in German and in Russian was taken from the whole corpus (Estonian and Livonian newspapers mixed; see the full code in `3_np_1880.ipynb`). After selection of pages, the words in texts were counted and, in case of Cyrillic script, transformed to modern orftography and lemmatized. Transformation to the modern orthography was made with `russpelling` Python module (see implementation in `scr/new_ofr.py`) and lemmatization performed with `pymystem3` (see `scr/lemmatization.ipynb`).

```{r, eval=FALSE, echo = TRUE}
# selection of pages and tokenization example:

# pages selection: 3000 pages in German and Russian
np_1880_langs <- np_1880_langs %>% 
    group_by(lang_cln) %>% 
    slice_sample(n = 3000) %>% 
    ungroup()

# tokenization
tokens <- np_1880_langs %>%
    unnest_tokens(input = text, output = word, token = "words")
top_tokens_1880 <- tokens %>% 
    group_by(PageID, word) %>% 
    count(sort = T) %>%
    ungroup()

stoplist <- c("что", "это", "быть", "тот", "который", "", "они", "для", "наш", "свой", "этот", "тот", "как", "все", "год","при", "час", "она", "есть", "чтобы", "давать", "один", "только", "время", "кто", "так", "огь", "уже", "ваш", "сто", "amp","под", "сие", "или", "день", "оть", "пта","январь", "февраль", "март", "апрель", "май", "июнь", "июль", "август", "сентябрь", "октябрь", "ноябрь", "декабрь","часть", "еще", "весь", "должный", "такой", "если", "себя", "сам", "окт", "том", "когда", "про", "прo", "копа", 
"кой", "сей", "мочь", "без",
"der", "die", "das", "dem", "den", "von", "des", "werden", "auf", "bei", "und", "mit", "vom", "zur", "uhr", "daß", "für", "zum", "sich", "wird", "welche", "ein", "eine", "eins", "groß", "ist", "mal", "aus", "oder", "nicht", "kop", "als","durch", "nach", "sie", "er", "es", "pfd", "bis", "per", "dito", "unter", "mai", "aus", "allen", "alle", "alles", "thlr", "diese", "dieses", "diesem", "diesen", "auch", "man", "sind", "ich", "nur", "einem", "wenn", "hat", "haben", "alt","einer", "sub", "worden", "belegene", "dess", "herrn", "herr", "über", "dieser", "aber", "nebst", "wie", "pol", "sammt", "jahr")

# fast cleaning: get rid of digits and smalles words (mostly noise)
top_tokens_1880 <- top_tokens_1880 %>% 
    filter(!str_detect(word, "^\\d+$|^\\W+$")) %>% 
    filter(!str_detect(word, "^\\w$|^\\w\\w$"))
    filter(!word %in% stoplist)

write.csv(top_tokens_1880, "tokens_1880.csv")

# resulting .csv file went through new_orf.py and lemmatization.ipynb scripts
```
  
For the model was used a selection of words longer than 3 characters, not in the stoplist and appeared at least 5 times in the corpus that results in 6745 unique words (35021 German and 52582 Russian tokens in total), 5789 documents.
An LDA model with 20 topics was build on the resulting data. 

As can be seen from the most probable words in each topic, the model easily distinguish German and Russian, e.g. uniquely-German topics are the 7th, 12th, etc. Content-wise topics in German and in Russian are very similar, containing mostly official news or advertisements. In both languages, the official news-related topics seems to be focused on local issues (naming places, governmental units such as magistrates and local courts), the advertisements are about sales, property, prices, etc.    

```{r, echo = F, include = T, fig.height=10, fig.width=8}
top_beta_1880 <- beta_1880 %>% 
  group_by(topic) %>% 
  slice_max(beta, n = 10) 

top_beta_1880 %>%
  ggplot(aes(x = beta, y = reorder_within(term, beta, topic), 
             fill = factor(topic))) + 
  geom_col(show.legend = FALSE) + 
  scale_y_reordered() + 
  facet_wrap(~ topic, scales = "free") + 
  scale_fill_manual(values = cols25(n = 20)) + 
  labs(x = "", y = "", 
       title = "Most probable words in topics")
```
  
The only topic that is significantly different from the rest is the 8th, which clearly belongs to ideological side of Russian politics: its most probable words are "Russian", "society", "church", "common", "language", "path", "committee", "orthodox", "region", "school". These are the keywords highly important for the discourse of Russification: it gives an imperialistic view on the citizens of the empire as one nation ("Russian") as opposed to minorities living in the governorates of Estonia and Livonia; as in the middle of the 19th century, the ideological support to Russification was given by the church (appearance of the word "orthodox"). The topic also refers to the language politics and education reforms in the region. All the 10 of these words does not appear as most probable in other topics.   
The question is: how much is this topic visible in the corpus? 

```{r, echo = F, include = T, out.width="50%", comment=F, warning=F, message=F}
gamma_1880 <- gamma_1880 %>% 
  mutate(title = str_remove_all(document, "\\d+|\\W")) %>% 
  mutate(year = str_replace_all(document, 
                                "\\w+(\\d{4})(\\d{2})(\\d{2})\\..*", "\\1")) %>% 
  mutate(month = str_replace_all(document, 
                                 "\\w+(\\d{4})(\\d{2})(\\d{2})\\..*", "\\2"))

gamma_years <- gamma_1880 %>% 
  group_by(year, topic) %>% 
  summarise(gamma_avg = mean(gamma))

gamma_summary <- gamma_1880 %>% 
  group_by(title, year, topic) %>% 
  summarise(gamma_avg = mean(gamma))

gamma_years %>% 
  ggplot(aes(x = year, y = gamma_avg, fill = factor(topic))) + 
  geom_col() + 
  scale_fill_manual(values = cols25(n = 20)) + 
  labs(x = "Year", y = "Avg topic probability", fill = "Topic")

```

If we look at the averaged topic distribution, the general assumption to have different sets of topics before and after 1885 is confirmed: the language difference is caught by the model. The presence of topics as the 16th shows that there were always some language mixing also before 1885. The topic 8 appeared in 1886 and takes the biggest part in 1888. However, there is a difference in topic distribution if we take the Estonian and Livonian newspapers separately: 

```{r, include = T, echo = F, fig.height=8, fig.width=5}
gamma_summary %>% 
  ggplot(aes(x = year, y = gamma_avg, fill = factor(topic))) + geom_col() + 
  facet_wrap(~title, scales = "free", ncol = 1) + 
  scale_fill_manual(values = cols25(n = 20)) + 
  labs(x = "", y = "", fill = "Topic")
```

According to the random sample used, the presence of several topics are exclusive for the newspaper of the Governorate of Estonia, including the "ideological" topic 8. Different sets of topics appeared in the two newspapers suppose that the language of news were different despite geographical and linguistic (predominantly German-language governorates) closeness. Difference in language use does not, however, explains why other ideological topic did not appear in Livonian governorate's newspaper at the same time as topic 8 appears at *"ekmteataja"*. One can suppose that there were probably another printed media in Riga to communicate the issue or that the Russification process was going differently in two regions: the problem may be investigated further using both distant and close reading techniques. 
  
---
### II
#### Estonian newspapers in 1917
  
Another period full of events is the year of two Russian revolutions. This is probably the only and last period of Russian-language printing in Estonia when several new newspapers appeared at the time to cover current events. New printed media appeared as an answer to great changes in political life, as it was said, for example, in center-oriented newspaper *The word of Revel"* ("Ревельское слово"):  
  
> "'The Word of Revel' is being born during the hard times of timelessness, in the twilight of sinister uncertainties, in the time of a nightmare suffocation. This defines the conditions of the very existence of the new print media, of the new broadcaster of the word which is alive, fresh, truthful, brave, and, as far as possible, free. <...> These days we cannot keep silence: our position obliges to speak about current needs, negative sides of the present times, which construct the society's evil <...>."     
  
```{r, echo = F, include=T, out.width="50%"}
# load data
beta_1917 <- read.csv("data/models/beta_1917_20.csv")
gamma_1917 <- read.csv("data/models/gamma_1917_20.csv")

# plot for new newspapers appeared??
meta <- read.csv("data/meta_rus_lang.csv")

meta %>% 
  filter(year.x > 1890 & year.x < 1920) %>% 
  mutate(title_cln = str_remove_all(docid, "\\d+")) %>% 
  group_by(year.x, title_cln) %>% 
  count() %>% 
  ungroup() %>% 
  select(-n) %>% 
  group_by(year.x) %>% 
  count() %>% 
  ggplot(aes(x = year.x, y = n)) + geom_col(fill = wes_palette("Chevalier1")[1]) + 
  labs(x = "Year", y = "Number of newspapers")

```

Below I would like to test similar topic modeling approach to see, if it can catch differences between some of the newspapers appeared in 1917 and well-established official newspaper of the Governorate of Estonia (the same discussed above). The hypothesis is that the newly appeared newspapers would speak about different topics than the official one.  
  
To test the hypothesis a topic model was build, taking into consideration _______ biggest newspapers of the 1917. Roughly these newspapers may be characterized as 1) the oldest official one (*The newspaper of the Governement of Estonia*), 2) Old Revel local newspaper *Revel News* ("Ревельские известия") 3) more center-oriented new local Tallinn newspapers: *The Word of Revel* ("Ревельское слово"), 4) revolution-inspired press of the social-democrats ("Известiя Ревельскаго совета рабочих и воинских депутатов", "Утро правды", "Звезда") and soldiers and sailors ("Свободное слово солдата и матроса").  

```{r, echo = F, include = T}
t <- tibble(PageID = gamma_1917$document) %>% distinct
t <- left_join(t, meta_1917, by = "PageID")
t %>% 
  filter(!str_detect(title, "Наше слово")) %>% 
  distinct(title, firstyear, lastyear) %>% 
  kbl(caption = "Newspapers of the 1917 in the corpus",
      col.names = c("First year of issue", "Last year", "Title")) %>% 
  kable_classic(full_width = F, lightable_options = c("striped"), latex_options="scale_down")

rm(meta, t)
```

The data includes 3228 pages from which the words longer than 2 characters and not included in the stoplist for Russian were taken (8174 unique words / lemmata). The LDA topic model build for 20 topics which are in this case monolongual.  

From the topic distribution on the timescale, the most different from the others is the oldest official newspaper (*ekmteataja*), which seems to discuss local themes in somewhat of archaic discourse: its most probable topic 20 refers to local issues such are "city" "Revel" "court", etc., more or less in the same manner as in the 1880s; topics 7 and 17 appears to be more contemporary discussing "counsil", "gouvernment", "minister"[s] but it seems none of other sources uses this vocabulary in the same way.  

For instance, the most visible topic for the left-wing *The Morning of the Truth* ("Утро правды"), *Tallinna Tööliste ja Sõjawäeliste Saadikute Nõukogu Teataja* and 

```{r}
gamma_1917 <- gamma_1917 %>% 
  mutate(title = str_remove_all(document, "\\d+|\\W")) %>% 
  mutate(year = str_replace_all(document, 
                                "\\w+(\\d{4})(\\d{2})(\\d{2})\\..*", "\\1")) %>% 
  mutate(month = str_replace_all(document, 
                                 "\\w+(\\d{4})(\\d{2})(\\d{2})\\..*", "\\2"))

gamma_month_title <- gamma_1917 %>% 
  group_by(title, month, topic) %>% 
  summarise(gamma_avg = mean(gamma))

gamma_month <- gamma_1917 %>% 
  group_by(month, topic) %>% 
  summarise(gamma_avg = mean(gamma))

gamma_month_title %>% 
  filter(title != "nasheslovo") %>%  # includes only 2 months
  ggplot(aes(x = month, y = gamma_avg, fill = factor(topic))) + 
  geom_col() + 
  scale_fill_manual(values = cols25(n = 20)) + 
  facet_wrap(~ title, scales = "free", ncol = 2)
```

```{r, include = F, echo = F}
top_beta_1917 <- beta_1917 %>% 
  group_by(topic) %>% 
  slice_max(beta, n = 10) 

top_beta_1917 %>%
  ggplot(aes(x = beta, y = reorder_within(term, beta, topic), 
             fill = factor(topic))) + 
  geom_col(show.legend = FALSE) + 
  scale_y_reordered() + 
  facet_wrap(~ topic, scales = "free") + 
  scale_fill_manual(values = cols25(20))
```




```{r}
```